---
title: "Lab 03"
author: "Jennifer Lin"
date: "jenniferyjlin@berkeley.edu"
output:
  html_document: default
  pdf_document: default
---
## 1
#### Come up with a system of equations with 3 unknowns and 3 equations 
#### (in contrast to the example with 2 unknowns and 2 equations) and solve it.
```{r}
A <- rbind(c(1,5,7), c(8,3,8), c(5,7,2))
b <- c(9,3,1)
x <- solve(A,b)
print(x)
```

## 2
#### Solve the following differential equation exactly as was done in the example. 
#### You may do it by hand or with code:
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>



## 3
#### Write code that models the predator-prey relationship (Lotka-Volterra model).
```{r}
func3 <- function(time3, state3, parameters3){
  # c in the equation in the lab_03 paper is actually k
  dV = parameters3['r']*state3['V']-parameters3['k']*state3['V']*state3['P']
  dP = parameters3['e']*parameters3['k']*state3['V']*state3['P'] - parameters3['d']*state3['P']
  # return(c(list(dV),list(dP)))
  return(list(c(dV,dP)))
}
```


## 4
#### Solve it with the following parameters and plot the results: r = 1.8, k = 0.4, e = 0.12, d = 1
```{r}
time3 <- seq(from=0, to=50, by=0.01)
parameters3 <- c(r=1.8, k=0.4, e=0.12, d=1)
state3 <- c(V=1, P=3)  # just try some numbers

# install.packages('deSolve',repos='http://probability.ca/cran')
library(deSolve)
output3 <- ode(y=state3, times=time3, func=func3, parms=parameters3)

plot(output3)
library(ggplot2)
q3_V <- ggplot(data=as.data.frame(output3),aes(time3,V)) + geom_point()
print(q3_V)
q3_P <- ggplot(data=as.data.frame(output3),aes(time3,P)) + geom_point()
print(q3_P)
```

