---
title: "Lab 08"
author: "Jennifer Lin"
date: "jenniferyjlin@berkeley.edu"
output:
  html_document: default
  pdf_document: default
---
<p>&nbsp;</p>

## 1
#### Read in the two sequences using the function provided and calculate the number of pairwise differences d.
```{r}
# Read in the two sequences
x <- read.table('lab_08_data.txt',sep = '\t')
seq1 <- strsplit(as.vector(x$V1[2]),split = '')[[1]]
seq2 <- strsplit(as.vector(x$V1[3]),split = '')[[1]]
# Calculate the number of pairwise differences d
d_result <- 0
S_result <- length(seq1)
for (i in 1:S_result){
  if(seq1[i] != seq2[i]){
    d_result = d_result + 1
  }
}
print(S_result) # total length of the sequence
print(d_result) # pairwise differences d
```

## 2
#### Implement a likelihood function based on the Jukes & Cantor model. 
#### The likelihood function is parameterized in terms of λ which is the genetic distance.
#### λ is proportional to the mutation rate and the amount of time the two species have been diverging from each other. 
#### The type of genetic distance is fundamental in many phylogenetic studies and other studies comparing DNA sequences.
```{r}
# JC_likelihood
# ( 1/4 + 3*(exp(-4*parameter_lambda/3))/4 ) ^ ( S-d )  *  ( 1/4 - 1*(exp(-4*parameter_lambda/3))/4 ) ^ ( d )
```

## 3
#### Create an R function that calculates the logarithm of the likelihood. 
#### This function requires 3 arguments: S,d, and λ.
```{r}
JC_likelihood_log <- function(S, d, parameter_lambda){
  return(log(  ( 1/4 + 3*(exp(-4*parameter_lambda/3))/4 ) ^ ( S-d )  *  ( 1/4 - 1*(exp(-4*parameter_lambda/3))/4 ) ^ ( d )  ))
}
```

## 4
#### Use the optim function in R to optimize the log-likelihood function for λ using the value of S and d, from the two sequences.
#### Does this value match the one observed in the plot?
```{r}
# Optim
optim_4 <- optim(par = c(1), fn = JC_likelihood_log, S = S_result, d = d_result)
print(optim_4$par[1]) # parameter_lambda
# Plot
parameter_lambda_seq <- seq(-2,2,by=0.1)
log_likelihood <- rep(NA,length(parameter_lambda_seq))
likelihood_grid_lambda <- as.data.frame(cbind(parameter_lambda_seq, log_likelihood))
for (i in 1:length(likelihood_grid_lambda$parameter_lambda_seq)){
  likelihood_grid_lambda$log_likelihood[i] <- JC_likelihood_log(S_result, d_result, likelihood_grid_lambda$parameter_lambda_seq[i])
}
plot(y=likelihood_grid_lambda$log_likelihood, x=likelihood_grid_lambda$parameter_lambda_seq)
# Yes, the result value from "optim" matches the one observed in the plot (lambda = ~1.4).
# (the minimum of log likelihood = the maximum of negative log likelihood)
```

## 5
#### Transitions (changes between A and G and between C and T) tends to occur at a higher rate than transversions (all other changes). 
#### Calculate the number of transition ds and the number of transversions dv from the two sequences.
#### The Kimura two-parameter model, is specified in terms of two parameters α and β. 
#### These two parameters can be interpreted as the rate of transitions and tranversions per unit time, respectively, multiplied by the total amount of time. 
#### Make a function in R that calculates the logarithm of this likelihood. It should take 5 arguments: S, ds, dv, α, and β
```{r}
K_likelihood_log <- function(S, ds, dv, parameters){
  parameter_alpha <- parameters[1]
  parameter_beta <- parameters[2]
  p1 <- ( 1/4 + 1*(exp(-4*parameter_beta))/4 + 1*(exp(-2*(parameter_alpha+parameter_beta)))/2 ) ^ ( S-dv-ds )
  p2 <- ( 1/4 + 1*(exp(-4*parameter_beta))/4 - 1*(exp(-2*(parameter_alpha+parameter_beta)))/2) ^ ( ds )
  p3 <- ( 1/4 - 1*(exp(-4*parameter_beta))/4 ) ^ ( dv )
  return(log(p1*p2*p3))
}
```

## 6
#### Plot the log likelihood function for the values of S, ds, and dv you obtained from the two sequences as a function of α and β in a contour plot.
#### Approximately where is the maximum likelihood estimate? 
#### For contour plots, please use the plotly package, please refer to this link.
```{r}
# calculate num_transversion and num_transition
num_transition <- 0
num_transversion <- 0
for (i in 1:length(seq1)){if (seq1[i] != seq2[i]){
  if ((seq1[i] == 'A' & seq2[i] == 'G') | (seq1[i] == 'G' & seq2[i] == 'A') | (seq1[i] == 'T' & seq2[i] == 'C')| (seq1[i] == 'C' & seq2[i] == 'T')){
    num_transition <- num_transition + 1}
  if ((seq1[i] == 'A' & seq2[i] == 'C') | (seq1[i] == 'C' & seq2[i] == 'A') | (seq1[i] == 'G' & seq2[i] == 'T') | (seq1[i] == 'T' & seq2[i] == 'G')){
    num_transversion <- num_transversion + 1}}}
# create a contour plot
library(plotly)
parameter_alpha_seq <- seq(-2,2,by=0.1)
parameter_beta_seq <- seq(-2,2,by=0.1)
likelihood_grid <- matrix(NA, length(parameter_alpha_seq),length(parameter_beta_seq))
for (i in 1:length(parameter_alpha_seq)){
  for (j in 1:length(parameter_beta_seq)){
    likelihood_grid[i,j] <- K_likelihood_log(S=S_result, ds=num_transition, dv=num_transversion, parameters=c(parameter_alpha_seq[i],parameter_beta_seq[j]))
  }
}
likelihood_grid[which(!is.finite(likelihood_grid))] <- NA # Inf is replaced with NA in order to prevent the Error "Wasn't able to determine range of domain"
fig <- plot_ly(z = likelihood_grid, type = 'contour', x = parameter_alpha_seq, y = parameter_beta_seq, colorscale = 'Viridis')
fig
```

## 7
#### Use the optim function in R to optimize this log-likelihood function for α and β using the value of S, ds, and dv from the two sequences.
#### Does this value match the one observed in the plot?
```{r}
optim_7 <- optim(par = c(0,0.5), fn = K_likelihood_log, S = S_result, ds = num_transition, dv = num_transversion)
print(optim_7$par[1]) # parameter_alpha
print(optim_7$par[2]) # parameter_beta
# Yes, the result value from "optim" roughly matches the one observed in the plot.
# (the minimum of log likelihood = the maximum of negative log likelihood)
# According to the contour plot, values in the right upper part are all quite low (we are searching for minimum in this plot).
# Depends on the initial value of the optim function, we may have different results.
# However, these results are all very close to the right upper part of the contour plot.
```